A Formalization of Typed Lua

### 2015-08-20 doc/papers/dls15/paper.tex
Fixing some typos for DSL'15 camera ready

\begin{figure}[t]
{\scriptsize
$$
\begin{array}{rlr}
F ::= & & \textsc{first-level types:}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{\overline{F:V}\}_{unique|open|fixed|closed} & \textit{table types}\\
& |\; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
&  \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \Boolean \ |\  \Integer \ |\ \Number\ |\ \String &\\\\
%\multicolumn{3}{c}{}\\
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; F \times P & \textit{pair types}\\\\
\end{array}
$$
}
\caption{Type Language}
\label{fig:typelang}
\end{figure}

\begin{figure}[t]
{\footnotesize
$$
\begin{array}{rlr}
T ::= & & \textsc{{\small expression types:}}\\
& \;\; F & \\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
E ::= & & \textsc{{\small expression list types:}}\\
& \;\; T{*} & \\
& | \; T \times E & \\
%K ::= & & \textsc{{\small key types:}}\\
%& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\
R ::= & & \textsc{\small{filter result type:}}\\
& \;\; \mathbf{void} & \textit{void type}\\
& |\; F
\end{array}
$$
}
\caption{Special Types}
\label{fig:special}
\end{figure}

@@

The type $\Any$ is the dynamic type, and allows
interaction between statically and dynamically typed
portions of a program. It is neither a top nor a bottom
type, although it can work as either, as a value of
any type can be coerced to $\Any$ and vice-versa.
Gradual type systems augment these coercions with
information that can be used to blame the dynamically
typed of the program for type errors that may occur in the
statically typed part. Currently Typed Lua only 
warns the programmer that coercions to and from $\Any$
may be unsafe.

@@

Figure~\ref{fig:special} presents some types that, while
they may not be present in type annotations, play some
special role in the type system. {\em Expression types}
denote the type of expressions in the type system, which,
in addition to first-level types, can also be
{\em filter types}, which are the result of flow typing,
and {\em projection types}, which are the result of
decomposing a union of tuples. {\em Expression list types}
denote the type of lists of expressions. {\em Key types}
and {\em value types} have also been mentioned above. Finally, {\em filter result types}
are also used in flow typing; the {\bf void} type is
a type with no values.

@@

Rule {\sc S-TableUtoC} for subtyping between {\em unique} and {\em closed} table types is more permissive, as the typing rules of Section~\ref{sec:tables} guarantee that a {\em unique}
table type can only appear on the left side of a subtyping
judgment if the reference to the table with the {\em unique}
tag is being destroyed. This rule allows width subtyping, depth subtyping on any field, not only immutable fields, as well as the addition of new fields and the merging of existing fields, as long as the types of the values of new and merged fields include $\Nil$.

@@

We will revisit the subtyping rules among table types on
Section~\ref{sec:tables}.

@@

The dynamic type $\Any$ is only a subtype of itself and of
$\Value$, and no type except $\Any$ is a subtype of $\Any$.
The type $\Any$ interacts with other types through the
{\em consistency} and {\em consistent-subtyping} 
relations~\cite{siek2007objects}. We use the symbol $\sim$ for
the consistency, and $\lesssim$ for consistent-subtyping.

The type $\Any$ is trivially consistent with all types, and all types
are consistent with $\Any$. A type is also consistent with
itself. Two union types $F_1 \cup F_2$ and $F_3 \cup F_4$ are
consistent if either $F_1 \sim F_3$ and $F_2 \sim F_4$ or
$F_1 \sim F_4$ and $F_2 \sim F_3$. Two function types
$S_1 \rightarrow S_2$ and $S_3 \rightarrow S_4$ are consistent
if and only if $S_1 \sim S_3$ and $S_2 \sim S_4$. Two table types are consistent if and only if they have the same key types, and the value types of each key type are consistent. Two recursive types are consistent if and only if their contents are consistent assuming that their type variables are consistent. Unions of tuple types follow the same rules as unions of first-level types. Variadic types are consistent if and only if their base types are consistent. 
Finally, pair types $F_1 \times P_1$ and $F_2 \times P_2$ are
consistent if and only if $F_1 \sim F_2$ and $P_1 \sim P_2$.

@@

\begin{figure*}[t]
{\footnotesize
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-Constructor}\\
\dfrac{\begin{array}{c}
	\overline{\env_v, \penv \vdash e_1: F_1, \env_f}
	\;\;\;
\overline{\env_1, \penv \vdash e_2: F_2, \env_v}	\\
	T = \{\overline{F_1{:}vt(F_1,F_2)}\}_{unique} \;\;\;
	\env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
	\end{array}}
{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-NewFieldUnique}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{ \overline{F:V} \}_{unique}\\
	\env_{1}, \penv \vdash e:F^{new}, \env_{2} \;\;\;
	\overline{F^{new} \not\lesssim F} \;\;\;
	V^{new} = vt(F^{new}, V)
	\end{array}}
{\begin{array}{l}
	\env_{1}, \penv \vdash id[e] {<}V{>}:V^{new},
	\env_{2}[id \mapsto \{ \overline{F{:}V}, F^{new}{:}V^{new}\}_{unique}]
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-IdRead}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id:close(F), \env[id \mapsto open(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-IdWrite}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id_{l}:F, \env}
\end{array}
\begin{array}{c}
\mylabel{T-IndexIdRead}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{\overline{F:V}\} \;\;\;
	\env_{1}, \penv \vdash e_{2}:F^e, \env_{2} \;\;\;
	 F^e \lesssim F_{i}
	\end{array}}
{\env_{1}, \penv \vdash id[e_{2}]:rconst(V_{i}), \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IndexExpRead}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{1}:\{\overline{F:V}\}, \env_{2}\\
	\env_{2}, \penv \vdash e_{2}:F^e, \env_{3} \;\;\;
	 F^e \lesssim F_{i}
	\end{array}}
{\env_{1}, \penv \vdash e_{1}[e_{2}]:rconst(V_{i}), \env_{3}}
\end{array}
\begin{array}{c}
\mylabel{T-CoerceClosed}\\
\dfrac{\env(id) \subtype F \;\;\; tag(F,closed)}
{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto reopen(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-CoerceFixed}\\
\dfrac{\env(id) \subtype F \;\;\; tag(F,fixed)}
{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto F]}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-Function}\\
\dfrac{\begin{array}{c}
	closeall(\env_{1})[\overline{id \mapsto F}], \penv[\ret \mapsto S] \vdash s, \env_{2}\\
	\env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}F}){:}S \; s)) \\
	\env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}F}){:}S \; s))
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}F}){:}S \; s:F_1 \times \ldots F_n \times \Void \rightarrow S, \env_{4}}
\end{array}
\begin{array}{c}
\mylabel{T-Local}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash el:P, \env_{2} \\
	P \lesssim F_1 \times \ldots \times F_n \times \Value{*} \\
	\env_{2}[\overline{id \mapsto F}], \penv \vdash s, \env_{3}
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s, (\env_{3} - \{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
\end{array}
\end{array}
\]
}
\caption{Typing Rules}
\label{fig:tablerules}
\end{figure*}

@@

The main differences between Core Typed Lua and the concrete
syntax are:
\begin{itemize}
\item lack of labels and goto statements, as they are incompatible
with the {\em flow typing} of Typed Lua's type system;
\item local variable declarations have an explicit block,
instead of being just statements;
\item no {\bf elseif} clause in {\bf if} statements;
\item no {\bf repeat} and {\bf for} loops, just {\bf while};
\item the only arithmetic operator is $+$, the only relational
operators are $==$ and $<$, and the only bitwise operator
is $\&$, as all other operators have typing rules identical to
one of these;
\item type coercion for evolution of table types is explicit,
with two coercion terms, the l-value $id[k] \; {<}V{>}$ for adding a field to 
a table through assignment, and the r-value ${<}F{>} \;id$ for reinterpreting
a table type to a compatible type that is expected by the context;
\item all function declarations are anonymous, so there is a
statement for recursive declarations;
\item function or method applications that throw away all
results or all but the first result are explicitly annotated
as such, instead this being implicit in their syntactic context;
\item identifiers that appear in l-value position are explicitly
annotated as such with an $l$ subscript.
\end{itemize}

@@

\begin{figure}
	{\footnotesize
	\[
	\begin{array}{rcl}
	wf(\{\overline{F:V}\}_{u|o|f|c}) & = &
	 \forall i \; ((\nexists j \; i \not= j \,\wedge\, F_{i} \lesssim F_{j}) \,\wedge \\
	 & &  wf(V_{i}) \,\wedge\, \lnot tag(V_{i},unique) \,\wedge\\
	 & &  \lnot tag(V_{i},open))\\
	wf({\bf const}\; F) & = & wf(F) \\
	wf(F_1 \cup F_2) & = & wf(F_1) \,\wedge\, wf(F_2) \\
	wf(\mu x.F) & = & wf(F) \\
	wf(S_1 \rightarrow S_2) & = & wf(S_1) \,\wedge\, wf(S_2) \\
	wf(S_1 \sqcup S_2) & = & wf(S_1) \,\wedge\, wf(S_2)\\
	wf(F{*}) &=& wf(F) \\
	wf(F \times P) &=& wf(F) \,\wedge\,wf(P)\\
	wf(F) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}\\\\
	tag(F_1 \cup F_2, t) & = & tag(F_1, t) \,\vee\, tag(F_2,t) \\
	tag(\{\overline{F:V}\}_t, t) & = & \top\\
	\multicolumn{3}{c}{
		\;\;\;\;\mathrm{other\;recusive\;cases\;are\;similar,\;other\;base\;cases\;are\;}\bot}\\
\\\\
    vt(L, V) & = & fix(V) \\
    vt(F_1, F_2) & = & nil(fix(F_2))\\
    vt(F_1, {\bf const}\;F_2) & = & {\bf const}\;nil(fix(F_2))\\\\
    nil(T) & = & T \;\;\; \mathrm{if}\;\Nil \lesssim T\\
    nil(T) & = & T \cup \Nil \;\;\; \mathrm{otherwise}\\\\
	fix(\{\overline{F:V}\}_{unique|open}) & = &
	\{\overline{F:V}\}_{fixed}\\
	\multicolumn{3}{c}{\mathrm{other\;cases\;by\;straightforward\;structural\;recursion}}
	\end{array}
	\]}
	\caption{Auxiliary Functions}
\label{fig:funcs}
\end{figure}

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
At the moment, local variables are the only kind of reference that the system
can track, so this restriction is
necessary. This limitation of the type system is the reason the {\sc T-Constructor} rule uses the auxiliary function $fix$
(in the definition of $vt$)
 to change any {\em open} and {\em unique} table types used in the field initializer to {\em fixed}. As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{fixed}\}_{unique}$.

There are two ways the type of a {\em unique} and {\em open} tables can change in Typed
Lua. The first way is by assignment to a non-existing
field. Let us go back to the example shown in Figure~\ref{fig:newpoint}. Assuming that the type of the
$x$ and $y$ is $\Number$, we can translate the first three lines
of function {\tt new\_point} to the abstract syntax of Figure~\ref{fig:syntax}:
{\footnotesize
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; point: \{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$point[``x"] \; {<}\Number{>} = x;\;
point[``y"] \; {<}\Number{>} = y$}
\end{tabular}
\end{center}
}

@@

Open tables have a rule {\sc T-NewFieldOpen} that is not shown
in Figure~\ref{fig:tablerules} due to brevity. The only thing that
changes from one rule to the other is the tag in the type of $id$.
There are analogous rules for adding methods to a {\em unique}
and {\em open tables} as a side-effect of typechecking a method declaration.

@@

Rule {\sc T-IdRead} is responsible for tracking
aliasing. Its auxiliary function $close$ is similar to
function $fix$ defined in Figure~\ref{fig:funcs},
except that it changes the tag of {\em unique} or
{\em open} tables to {\em closed}. Its purpose
is to construct a {\em closed} type for the alias,
if necessary. Its auxiliary function $open$ is similar
to $fix$ and $close$, except that it changes the type of
{\em unique} tables to {\em open}. Its purpose
is to track the fact that a $id$ now has an alias,
if necessary.

Consider the following example, where we make an alias
of a {\em unique} table:
{\footnotesize
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{closed} = a \; \mathbf{in}\;\;a[``x"] \; {<}\String{>} = ``foo";$}
\end{tabular}
\end{center}
}

@@

If we did not change the tag on the type produced by $a$
we could break type safety, as we could declare $b$ with
tag {\em unique} or {\em open} and add fields that would
conflict with the type of the fields on $a$, but at
runtime both reference the same table.

@@

A second way of changing a table type is by a 
{\em coercion expression}. Coercion expressions let
a {\em unique} or {\em open} table be used in a context
that expects a table type with a different shape.
Coercion expressions always create an alias, so
the type produced by the coercion is always
{\em closed} or {\em fixed}. If the coercion is to a
{\em closed} table type the type of the table changes
to an {\em open} type with the same shape, but if
the coercion is to a {\em fixed} type the table has
to assume the same type.

@@

In the example below, we use a coercion expression to
widen the type of a field in a {\em unique}
table, and then add another field, to show that the
table is now {\em open}:
{\footnotesize
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$a[``x"] \; {<}\String{>} = ``foo";\;
a[``y"] \; {<}\String{>} = ``bar";$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
& & \multicolumn{2}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{closed}{>} \; a \; \mathbf{in}$}\\
& & & \multicolumn{1}{l}{$a[``z"] \; {<}\Integer{>} = 1$}
\end{tabular}
\end{center}
}

>>>>

\begin{figure}[t]
{\scriptsize
$$
\begin{array}{rlr}
F ::= & & \textsc{first-level types:}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{\overline{F:V}\}_{unique|open|fixed|closed} & \textit{table types}\\
& |\; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\tiny literal types:}}\\
&  \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\tiny base types:}}\\
& \Boolean \ |\  \Integer \ |\ \Number\ |\ \String &\\\\
%\multicolumn{3}{c}{}\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\tiny value types:}}\\
& \;\; F \; | \; \Const \; F &\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\tiny tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; F \times P & \textit{pair types}\\\\
\end{array}
$$
}
\caption{Type Language}
\label{fig:typelang}
\end{figure}

\begin{figure}[t]
{\footnotesize
$$
\begin{array}{rlr}
T ::= & & \textsc{expression types:}\\
& \;\; F & \textit{first-level types}\\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
E ::= & & \textsc{expression list types:}\\
& \;\; T{*} & \textit{variadic types}\\
& | \; T \times E & \textit{pair types}\\
%K ::= & & \textsc{{\small key types:}}\\
%& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
R ::= & & \textsc{filter result type:}\\
& \;\; \mathbf{void} & \textit{void type}\\
& |\; F & \textit{first-level types}
\end{array}
$$
}
\caption{Special Types}
\label{fig:special}
\end{figure}

@@

The type $\Any$ is the dynamic type, and allows
interaction between statically and dynamically typed
portions of a program. It is neither a top nor a bottom
type, although it can work as either, as a value of
any type can be coerced to $\Any$ and vice-versa.
Gradual type systems augment these coercions with
information that can be used to blame the dynamically
typed part of the program for type errors that may occur in the
statically typed part. Currently Typed Lua only
warns the programmer that coercions to and from $\Any$
may be unsafe.

@@

Figure~\ref{fig:special} presents some types that, while
they may not be present in type annotations, play some
special role in the type system. {\em Expression types}
denote the type of expressions in the type system, which,
in addition to first-level types, can also be
{\em filter types}, which are the result of flow typing,
and {\em projection types}, which are the result of
decomposing a union of tuples. {\em Expression list types}
denote the type of lists of expressions.
Finally, {\em filter result types}
are also used in flow typing; the {\bf void} type is
a type with no values.

@@

Rule {\sc S-TableUtoC} for subtyping between {\em unique} and {\em closed} table types is more permissive, as the typing rules of Section~\ref{sec:tables} guarantee that an {\em unique}
table type can only appear on the left side of a subtyping
judgment if the reference to the table with the {\em unique}
tag is being destroyed. This rule allows width subtyping, depth subtyping on any field, not only immutable fields, as well as the addition of new fields and the merging of existing fields, as long as the types of the values of new and merged fields include $\Nil$.

@@

We will revisit the subtyping rules among table types in
Section~\ref{sec:tables}.

@@

The dynamic type $\Any$ is only a subtype of itself and of
$\Value$, and no type except $\Any$ is a subtype of $\Any$.
The type $\Any$ interacts with other types through the
{\em consistency} and {\em consistent-subtyping}
relations~\cite{siek2007objects}. We use the symbol $\sim$ for
consistency, and $\lesssim$ for consistent-subtyping.

The type $\Any$ is trivially consistent with all types, and all types
are consistent with $\Any$. A type is also consistent with
itself. Two union types $F_1 \cup F_2$ and $F_3 \cup F_4$ are
consistent if either $F_1 \sim F_3$ and $F_2 \sim F_4$ or
$F_1 \sim F_4$ and $F_2 \sim F_3$. Two function types
$S_1 \rightarrow S_2$ and $S_3 \rightarrow S_4$ are consistent
if and only if $S_3 \sim S_1$ and $S_2 \sim S_4$. Two table types are consistent if and only if they have the same key types, and the value types of each key type are consistent. Two recursive types are consistent if and only if their contents are consistent assuming that their type variables are consistent. Unions of tuple types follow the same rules as unions of first-level types. Variadic types are consistent if and only if their base types are consistent. 
Finally, pair types $F_1 \times P_1$ and $F_2 \times P_2$ are
consistent if and only if $F_1 \sim F_2$ and $P_1 \sim P_2$.

@@

\begin{figure*}[t]
{\footnotesize
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-Constructor}\\
\dfrac{\begin{array}{c}
	\overline{\env_v, \penv \vdash e_1: F_1, \env_f}
	\;\;\;
\overline{\env_1, \penv \vdash e_2: F_2, \env_v}	\\
	T = \{\overline{F_1{:}vt(F_1,F_2)}\}_{unique} \;\;\;
	\env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
	\end{array}}
{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-NewFieldUnique}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{ \overline{F:V} \}_{unique}\\
	\env_{1}, \penv \vdash e:F^{new}, \env_{2} \;\;\;
	\overline{F^{new} \not\lesssim F} \;\;\;
	V^{new} = vt(F^{new}, V)
	\end{array}}
{\begin{array}{l}
	\env_{1}, \penv \vdash id[e] {<}V{>}:V^{new},
	\env_{2}[id \mapsto \{ \overline{F{:}V}, F^{new}{:}V^{new}\}_{unique}]
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-IdRead}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id:close(F), \env[id \mapsto open(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-IdWrite}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id_{l}:F, \env}
\end{array}
\begin{array}{c}
\mylabel{T-IndexIdRead}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{\overline{F:V}\} \;\;\;
	\env_{1}, \penv \vdash e_{2}:F^e, \env_{2} \;\;\;
	 F^e \lesssim F_{i}
	\end{array}}
{\env_{1}, \penv \vdash id[e_{2}]:rconst(V_{i}), \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IndexExpRead}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{1}:\{\overline{F:V}\}, \env_{2}\\
	\env_{2}, \penv \vdash e_{2}:F^e, \env_{3} \;\;\;
	 F^e \lesssim F_{i}
	\end{array}}
{\env_{1}, \penv \vdash e_{1}[e_{2}]:rconst(V_{i}), \env_{3}}
\end{array}
\begin{array}{c}
\mylabel{T-CoerceClosed}\\
\dfrac{\env(id) \subtype F \;\;\; tag(F,closed)}
{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto reopen(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-CoerceFixed}\\
\dfrac{\env(id) \subtype F \;\;\; tag(F,fixed)}
{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto F]}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-Function}\\
\dfrac{\begin{array}{c}
	closeall(\env_{1})[\overline{id \mapsto F}], \penv[\ret \mapsto S] \vdash s, \env_{2}\\
	\env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}F}){:}S \; s)) \\
	\env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}F}){:}S \; s))
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}F}){:}S \; s:F_1 \times \ldots F_n \times \Void \rightarrow S, \env_{4}}
\end{array}
\begin{array}{c}
\mylabel{T-Local}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash el:P, \env_{2} \\
	P \lesssim F_1 \times \ldots \times F_n \times \Value{*} \\
	\env_{2}[\overline{id \mapsto F}], \penv \vdash s, \env_{3}
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s, (\env_{3} - \{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
\end{array}
\end{array}
\]
}
\caption{Typing Rules for Table Types}
\label{fig:tablerules}
\end{figure*}

@@

The main differences between Core Typed Lua and the concrete
syntax are:
\begin{itemize}
\item lack of labels and goto statements, as they are incompatible
with the {\em flow typing} of Typed Lua's type system;
\item local variable declarations have an explicit block,
instead of being just statements;
\item no {\bf elseif} clause in {\bf if} statements;
\item no {\bf repeat} and {\bf for} loops, just {\bf while};
\item the only arithmetic operator is $+$, the only relational
operators are $==$ and $<$, and the only bitwise operator
is $\&$, as all other operators have typing rules identical to
one of these;
\item type coercion for evolution of table types is explicit,
with two coercion terms, the l-value $id[e] \; {<}V{>}$ for adding a field to 
a table through assignment, and the r-value ${<}F{>} \;id$ for reinterpreting
a table type to a compatible type that is expected by the context;
\item all function declarations are anonymous, so there is a
statement for recursive declarations;
\item function or method applications that throw away all
results or all but the first result are explicitly annotated
as such, instead this being implicit in their syntactic context;
\item identifiers that appear in l-value position are explicitly
annotated as such with an $l$ subscript.
\end{itemize}

@@

\begin{figure}
	{\footnotesize
	\[
	\begin{array}{rcl}
	wf(\{\overline{F:V}\}_{u|o|f|c}) & = &
	 \forall i \; ((\nexists j \; i \not= j \,\wedge\, F_{i} \lesssim F_{j}) \,\wedge \\
	 & &  wf(V_{i}) \,\wedge\, \lnot tag(V_{i},unique) \,\wedge\\
	 & &  \lnot tag(V_{i},open))\\
	wf({\bf const}\; F) & = & wf(F) \\
	wf(F_1 \cup F_2) & = & wf(F_1) \,\wedge\, wf(F_2) \\
	wf(\mu x.F) & = & wf(F) \\
	wf(S_1 \rightarrow S_2) & = & wf(S_1) \,\wedge\, wf(S_2) \\
	wf(S_1 \sqcup S_2) & = & wf(S_1) \,\wedge\, wf(S_2)\\
	wf(F{*}) &=& wf(F) \\
	wf(F \times P) &=& wf(F) \,\wedge\,wf(P)\\
	wf(F) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}\\\\
	tag(F_1 \cup F_2, t) & = & tag(F_1, t) \,\vee\, tag(F_2,t) \\
	tag(\{\overline{F:V}\}_t, t) & = & \top\\
	\multicolumn{3}{c}{
		\;\;\;\;\mathrm{other\;recusive\;cases\;are\;similar,\;other\;base\;cases\;are\;}\bot}\\
\\\\
    vt(L, V) & = & fix(V) \\
    vt(F_1, F_2) & = & nil(fix(F_2))\\
    vt(F_1, {\bf const}\;F_2) & = & {\bf const}\;nil(fix(F_2))\\\\
    nil(T) & = & T \;\;\; \mathrm{if}\;\Nil \lesssim T\\
    nil(T) & = & T \cup \Nil \;\;\; \mathrm{otherwise}\\\\
	fix(\{\overline{F:V}\}_{unique|open}) & = &
	\{\overline{F:V}\}_{fixed}\\
	\multicolumn{3}{c}{\mathrm{other\;cases\;by\;straightforward\;structural\;recursion}}
	\end{array}
	\]}
	\caption{Auxiliary Functions for Table Types}
\label{fig:funcs}
\end{figure}

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
At the moment, local variables are the only kind of reference that the system
can track, so this restriction is
necessary. This limitation of the type system is the reason why the {\sc T-Constructor} rule uses the auxiliary function $fix$
(in the definition of $vt$)
 to change any {\em open} and {\em unique} table types used in the field initializer to {\em fixed}. As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{fixed}\}_{unique}$.

There are two ways the type of {\em unique} and {\em open} tables can change in Typed
Lua. The first way is by assignment to a non-existing
field. Let us go back to the example shown in Figure~\ref{fig:newpoint}. Assuming that the type of the
$x$ and $y$ is $\Number$, we can translate the first three lines
of function {\tt new\_point} to the abstract syntax of Figure~\ref{fig:syntax}:
{\footnotesize
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; point: \{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$point[``x"] \; {<}\Number{>} = x;\;
point[``y"] \; {<}\Number{>} = y$}
\end{tabular}
\end{center}
}

@@

Open tables have a rule {\sc T-NewFieldOpen} that is not shown
in Figure~\ref{fig:tablerules} due to brevity. The only thing that
changes from one rule to the other is the tag in the type of $id$.
There are analogous rules for adding methods to {\em unique}
and {\em open tables} as a side-effect of typechecking a method declaration.

@@

Rule {\sc T-IdRead} is responsible for tracking
aliasing. Its auxiliary function $close$ is similar to
function $fix$ defined in Figure~\ref{fig:funcs},
except that it changes the tag of {\em unique} or
{\em open} tables to {\em closed}. Its purpose
is to construct a {\em closed} type for the alias,
if necessary. Its auxiliary function $open$ is similar
to $fix$ and $close$, except that it changes the type of
{\em unique} tables to {\em open}. Its purpose
is to track the fact that an $id$ now has an alias,
if necessary.

Consider the following example, where we make an alias
of an {\em unique} table:
{\footnotesize
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{closed} = a \; \mathbf{in}\;\;a[``x"] \; {<}\String{>} = ``foo";$}
\end{tabular}
\end{center}
}

@@

If we did not change the tag on the type produced by $a$
we could break type safety, as we could declare $b$ with
tag {\em unique} or {\em open} and add fields that would
conflict with the type of the fields in $a$, but at
runtime both reference the same table.

@@

A second way of changing a table type is by a 
{\em coercion expression}. Coercion expressions let
an {\em unique} or {\em open} table be used in a context
that expects a table type with a different shape.
Coercion expressions always create an alias, so
the type produced by the coercion is always
{\em closed} or {\em fixed}. If the coercion is to a
{\em closed} table type the type of the table changes
to an {\em open} type with the same shape, but if
the coercion is to a {\em fixed} type the table has
to assume the same type.

@@

In the example below, we use a coercion expression to
widen the type of a field in an {\em unique}
table, and then add another field, to show that the
table is now {\em open}:
{\footnotesize
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$a[``x"] \; {<}\String{>} = ``foo";\;
a[``y"] \; {<}\String{>} = ``bar";$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
& & \multicolumn{2}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{closed}{>} \; a \; \mathbf{in}$}\\
& & & \multicolumn{1}{l}{$a[``z"] \; {<}\Integer{>} = 1$}
\end{tabular}
\end{center}
}

The type of $a$ by the time of the inner {\bf local} block
is $\{``x":\String, ``y":\String\}_{unique}$. This type is
a subtype of $\{``x":\String, ``y":\String\cup\Nil\}_{closed}$, as {\em unique} tables
have depth subtyping even in non-{\bf const} fields.
The coercion expression and the initialization of $b$
typecheck, and the type of $a$ changes to
$\{``x":\String, ``y":\String\cup\Nil\}_{open}$.
As the type is {\em open}, we can continue adding fields
to it, but are now restricted on future coercions. For
example, we will not be able to widen the type of $``x"$
to $\String \cup \Nil$, as that could break type safety
(by writing $\Nil$ to $``x"$ using reference $a$, and reading
$``x"$ using the alias $b$).
